openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2022-05-22 paper/main.tex
Text edits.

%% - \Abstract{} almost as usual
\Abstract{
One of the most popular techniques for visualizing large, high-dimensional data sets is $t$-distributed stochastic neighbor embedding (t-SNE). Recently, several extensions have been proposed to address scalability issues and the quality of the resulting visualizations. We introduce \opentsne, a modular \proglang{Python} library that implements the core t-SNE algorithm and its many extensions. The library is fast and can treat data sets containing millions of data points. We release \opentsne in open source and distribute it under the BSD-3-Clause License. Its source code is available at \url{https://github.com/pavlin-policar/openTSNE}.
}

@@

The ever-growing volumes of high-dimensional data sets in machine learning call for efficient dimensionality reduction techniques and their implementations to produce informative data visualizations. Popular approaches include principal component analysis, multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE has received much attention as it can address high volumes of data and reveal the underlying data structure. For instance, t-SNE is widely used in the bioinformatics community in areas such as single-cell transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human genetics~\citep{hirata2019genetic}, metagenomic assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\citep{sheth2019spatial}, and metabolomics~\citep{tkachev2019differences}. To visually explore the data structure, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, the expression of genes or the concentration of metabolites. The task of t-SNE is to embed the objects of interest within a low, preferably two-dimensional embedding. Reports on single-cell gene expression data, our running example, often start with an overview of the cell landscape and use t-SNE to embed high-dimensional expression profiles into a two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show two such embeddings.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}
We use \opentsne to generate three t-SNE embeddings and demonstrate recent theoretical advances. The data in \textbf{(a)} and \textbf{(b)} includes 44,808 mouse retinal cells (dots in the plots) that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}. The data in \textbf{(c)} additionally contains 27,499 expression profiles from mouse retinal cells from \citet{shekhar2016comprehensive}. \textbf{(a)} We construct t-SNE embedding following the parameter choices from the original publication by \citet{maaten2008visualizing}. The visualization shows no preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods. \textbf{(b)} A modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points and initialized so that the global layout is as meaningful as possible. Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types. The embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be used to place new samples into existing embeddings. The positions of new data points correspond to cell types from the reference atlas.
}
\end{figure*}

Despite its utility, t-SNE has often been criticized for its limited scalability, lack of global organization -- t-SNE identifies well-defined clusters that may be arbitrarily scattered throughout the embedding, and the absence of theoretically-founded methods to map new data into existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}. Most of these shortcomings have recently been addressed. \citet{linderman2019fast} developed FIt-SNE, an efficient approximation scheme that massively improves the scalability of t-SNE, achieving linear time complexity in the number of samples. \citet{kobak2019art} proposed several techniques to improve global cluster coherence, including estimating similarities with a mixture of Gaussian kernels. In our previous work, we introduced a principled approach for embedding new samples into existing visualizations~\citep{policar2021embedding}.

However, easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations. In this paper, we present \opentsne, our open-source \proglang{Python} implementation of t-SNE, complete with its recently proposed extensions. \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

$t$-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method that to aims to find a low-dimensional embedding
where neighborhoods are preserved. More formally, given a multi-dimensional data set
$\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \}
\in \mathbb{R}^D$ where $N$ is the number of data points in the data set, t-SNE
aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1,
\mathbf{y}_2, \dots, \mathbf{y}_N \right\} \in \mathbb{R}^d$ where $d \ll D$,
such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the
high-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and
$\mathbf{y}_j$ are also close. Since t-SNE is primarily used as a visualization
tool, $d$ is typically set to two. The similarity between two data points in the
high-dimensional space is defined using the Gaussian kernel
\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
\label{eq:p_ij}
\end{equation}
where $\mathcal{D}$ is some distance measure. The values $p_{j \mid i}$
are then symmetrized to
\begin{equation}
p_{ij} = \frac{p_{j \mid i} + p_{i \mid j}}{2N}.
\label{eq:symmetrize}
\end{equation}

@@

A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
which makes its application impractical to any reasonably-sized data set and
highlights the need for the development of efficient approximation schemes.
\citet{van2014accelerating} observed that the t-SNE gradient can be cast as an
N-body problem where data points represent particles that attract and repel each
other. The gradient from Eqn.~(\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep} \notag
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
)^{-1}$. We can view this equation as a particle simulation, where the
two terms represent the attractive and repulsive forces between individual
particles. Each term lends itself to efficient approximations, enabling us to
reduce the time complexity of t-SNE to $\mathcal{O}(N)$. This allows modern
t-SNE implementations to be leveraged for the visualization of data sets
containing up to millions of data points.

\subsubsection*{Attractive Forces}
\citet{van2014accelerating} observed that evaluating the attractive forces
between all pairs of data points is excessive, and that considering only a
handful of nearest neighbors at each point are sufficient to obtain a good
approximation. Therefore, instead of calculating all pairwise interactions, it
is sufficient to only find and evaluate the attraction to a certain number of
$k$ nearest neighbors. Using tree-based nearest-neighbor search methods,
\citet{van2014accelerating} reduced the time complexity to $\mathcal{O}(N \log
N)$. \citet{linderman2019fast} further realized that, qualitatively, embeddings
are visually indistinguishable when using only \textit{approximate} nearest
neighbors, further reducing time complexity to $\mathcal{O}(N)$.

@@

The t-SNE algorithm is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Therefore embeddings of new data
points need to be found through optimization
techniques~\citep{policar2021embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place
while new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \qquad
q_{j \mid i} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}, \notag
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}). Plugging these terms into
Eqn.~(\ref{eq:kl_divergence}), we obtain the following gradient
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}. \notag
\label{eq:gradient}
\end{equation}

@@

\item The optimization in t-SNE starts with initial positions for each data point in the embedding space. We can set the initial positions randomly or impose some initial global structure on the embedding with, for example, a PCA-based initialization. Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, The user can also specify their initialization by providing a \code{numpy.ndarray} object. By default, \opentsne opts for PCA-based initialization.

@@

While \opentsne allows for control at every step, most end-users often only need a quick and easy way to create informative visualizations of their data sets. \opentsne provides a high-level API that closely follows the style of \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used machine learning toolkit.

@@

Exaggeration can highlight transitions between cell states in developmental studies. Standard t-SNE often produces embeddings with clearly defined, discrete clusters. We can adjust the level of granularity and resolution of the clusters with several parameters, as shown in Fig.~\ref{fig:tasic}. However, discrete clusters are often undesired in developmental studies where cells' state is assumed to follow a continuous transition path. To this end, researchers have used other embedding techniques such as UMAP and ForceAtlas2 to better capture the continuity between cell states. Recently, \citet{bohm2020unifying} showed that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings that are markedly similar to UMAP and ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters. On the other hand, it is easier to trace the development when we increase the exaggeration factor from $1$ to $2$ to $4$ in Figs.~\ref{fig:cao}.b-c.

@@

Versatility, the ability to use and combine different optimization approaches to construct different embedding spaces, is another of \opentsne's core design principles. \citet{kobak2019art} recently provided several recommendations and tricks to obtain better and more meaningful t-SNE visualizations. These include multi-scale similarity kernels, perplexity annealing, and increasing exaggeration when working with massive data sets. \opentsne provides a flexible API to incorporate these improvements in just a few lines of code. Furthermore, \opentsne supports custom affinity models, enabling users to construct t-SNE embeddings on non-tabular relational data: the only requirement imposed by the affinity model is some notion of similarity between data points. Finally, \opentsne's comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.

@@

% General problem statement
One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\citep {becht2019dimensionality}. The culprit for slow response time stems from the optimization procedures and their specific implementation in popular open-source libraries. Until quite recently, most popular implementations of t-SNE were based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation scheme developed by \citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}(N \log N)$. Recently, however, \citet{linderman2019fast} developed a novel approximation scheme -- FIt-SNE -- which reduces the asymptotic time complexity to $\mathcal {O}(N)$, overcoming the difficulties of scaling t-SNE to large data sets.

% We benchmark Python
Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.6.2). Benchmarks were performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original t-SNE parameters, as some implementations do not allow for their modification. To ensure a realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse brain data set~\citep{xxx}, containing 1.3 million single-cell gene-expression profiles. We preprocess the data using the standard single-cell analysis pipeline~\citep{xxx} and extract the top 50 principal components. To generate benchmark data sets of different sample sizes, we subsample the data set five times at each sample size. In total, we run each implementation on 30 different data sets.

>>>>

%% - \Abstract{} almost as usual
\Abstract{
One of the most popular techniques for visualizing large, high-dimensional data sets is $t$-distributed stochastic neighbor embedding (t-SNE). Recently, several extensions have been proposed to address scalability issues and the quality of the resulting visualizations. We introduce \opentsne, a modular \proglang{Python} library that implements the core t-SNE algorithm and its many extensions. The library is faster than existing implementations and can compute projections of data sets containing millions of data points in minutes. We release \opentsne in open source and distribute it under the BSD-3-Clause License. Its source code is available at \url{https://github.com/pavlin-policar/openTSNE}.
}

@@

The ever-growing volumes of high-dimensional data sets in machine learning call for efficient dimensionality reduction techniques and their implementations to produce informative data visualizations. Popular approaches include principal component analysis, multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE has achieved widespread adoption as it can address high volumes of data and reveal the underlying data manifold. For instance, t-SNE is widely used in the bioinformatics community in areas such as single-cell transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human genetics~\citep{hirata2019genetic}, metagenomic assembly~\citep{beaulaurier2018metagenomic}, spatial organization of microbial communities~\citep{sheth2019spatial}, and metabolomics~\citep{tkachev2019differences}. To visually explore the data manifold, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, the expression of genes or the concentration of metabolites. The objective of t-SNE is to embed data points within a low-dimensional space, where the preservation of distances is rewarded in a non-linear fashion, thus resulting in a compact grouping of highly similar data points. Reports on single-cell gene expression data, where increasingly specific cell types are distinguished by a decreasing number of genes, 
often rely on t-SNE to embed high-dimensional expression profiles into a two-dimensional space and discover constituent cell types (Fig.~\ref{fig:macosko}.a).

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}
We use \opentsne to generate three t-SNE embeddings and demonstrate recent theoretical advances. The data in \textbf{(a)} and \textbf{(b)} include 44,808 mouse retinal cells that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}. The data in \textbf{(c)} additionally contains 27,499 expression profiles from mouse retinal cells from \citet{shekhar2016comprehensive}. \textbf{(a)} We construct t-SNE embedding following the parameter choices from the original publication by \citet{maaten2008visualizing}. The visualization shows no preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods. \textbf{(b)} A modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points and initialized so that the global layout is as meaningful as possible. Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types. The embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be used to place new samples into existing embeddings. The positions of new data points correspond to cell types from the reference atlas.
}
\end{figure*}

Despite its utility, t-SNE has often been criticized for its limited scalability, lack of global organization -- t-SNE identifies well-defined clusters that may be arbitrarily scattered throughout the embedding -- and the absence of theoretically-founded methods to map new data into existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}. Most of these shortcomings have recently been addressed. \citet{linderman2019fast} developed FIt-SNE, an efficient approximation that massively improves the scalability of t-SNE, achieving linear time complexity in the number of samples. \citet{kobak2019art} proposed several techniques to improve global cluster coherence (Fig. \ref{fig:macosko}.b), including estimating similarities with a mixture of Gaussian kernels. In our previous work, we introduced a principled approach for embedding new samples into existing visualizations (Fig.~\ref{fig:macosko}.c and~\citet{policar2021embedding}).

Easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations. In this paper, we present \opentsne, our open-source \proglang{Python} implementation of t-SNE, complete with its recently proposed extensions. \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

$t$-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method that to aims to find a low-dimensional embedding
where neighborhoods are preserved. Given a multi-dimensional data set with $N$ data points
$\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \}
\in \mathbb{R}^D$, t-SNE
aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1,
\mathbf{y}_2, \dots, \mathbf{y}_N \right\} \in \mathbb{R}^d$ where $d \ll D$,
such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the
high-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and
$\mathbf{y}_j$ are also close. Since t-SNE is primarily used as a visualization
tool, $d$ is typically set to two. The similarity between two data points in the
high-dimensional space is defined using the Gaussian kernel
\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
\label{eq:p_ij}
\end{equation}
where $\mathcal{D}$ is some distance measure. The values $p_{j \mid i}$
are then symmetrized to
\begin{equation}
p_{ij} = \frac{p_{j \mid i} + p_{i \mid j}}{2N}.
\label{eq:symmetrize}
\end{equation}

@@

Evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
which makes it impractical with large data sets and
calls for efficient approximation schemes.
\citet{van2014accelerating} observed that the t-SNE gradient can be cast as an
N-body problem where data points represent particles that attract and repel each
other. The gradient from Eqn.~(\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep} \notag
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
)^{-1}$. We can view this equation as a particle simulation, where the
two terms represent the attractive and repulsive forces between individual
particles. Each term lends itself to efficient approximations, enabling us to
reduce the time complexity of t-SNE to $\mathcal{O}(N)$. This allows modern
t-SNE implementations to be leveraged for the visualization of data sets
containing up to millions of data points.

\subsubsection*{Attractive Forces}
\citet{van2014accelerating} observed that evaluating the attractive forces
between all pairs of data points is excessive. In practice, it 
suffices to approximate the attraction forces against a small number of
$k$ nearest neighbors. Using tree-based nearest-neighbor search methods,
\citet{van2014accelerating} reduced the time complexity to $\mathcal{O}(N \log
N)$. \citet{linderman2019fast} noted that embeddings
are visually indistinguishable when using merely \textit{approximate} nearest
neighbors, further reducing time complexity to $\mathcal{O}(N)$.

@@

The t-SNE algorithm is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Embeddings of new data
points need to be found through optimization
schemes~\citep{policar2021embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place
while new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \qquad
q_{j \mid i} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}, \notag
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}). Plugging these terms into
Eqn.~(\ref{eq:kl_divergence}), we obtain the following gradient
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}. \notag
\label{eq:gradient}
\end{equation}

@@

\item The optimization in t-SNE starts with initial positions for each data point in the embedding space. We can set the initial positions randomly or impose an initial global structure on the embedding with, for example, a PCA-based initialization. Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, The user can also specify their initialization by providing a \code{numpy.ndarray} object. By default, \opentsne opts for PCA-based initialization.

@@

Most end-users often only need a quick and easy way to create informative visualizations of their data sets. \opentsne provides a high-level API that closely follows the style of \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used machine learning toolkit.

@@

Exaggeration can highlight transitions between cell stages in developmental studies. Standard t-SNE often produces embeddings with clearly defined, discrete clusters. We can adjust the level of granularity and resolution of the clusters with several parameters, as shown in Fig.~\ref{fig:tasic}. However, discrete clusters are often undesired in developmental studies where cells' stage is assumed to follow a continuous transition path. To this end, researchers have used other embedding techniques such as UMAP and ForceAtlas2 to better capture the continuity between cell stages. Recently, \citet{bohm2020unifying} showed that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings that are markedly similar to UMAP and ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters. On the other hand, it is easier to trace the development when we increase the exaggeration factor from $1$ to $2$ to $4$ in Figs.~\ref{fig:cao}.b-c.

@@

The ability to use and combine different optimization approaches to construct different embedding spaces is another of \opentsne's core design principles. \citet{kobak2019art} recently provided several recommendations and tricks to obtain better and more meaningful t-SNE visualizations. These include multi-scale similarity kernels, perplexity annealing, and increasing exaggeration when working with massive data sets. \opentsne provides a flexible API to incorporate these improvements in just a few lines of code. Furthermore, \opentsne supports custom affinity models, enabling users to construct t-SNE embeddings on non-tabular relational data: the only requirement imposed by the affinity model is a notion of similarity between data points. Finally, \opentsne's comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.

@@

% General problem statement
One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\citep {becht2019dimensionality}. Slow response times stem from the optimization procedures and their specific implementation in popular open-source libraries. Most implementations of t-SNE have been based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation with asymptotic time complexity $\mathcal{O}(N \log N)$~\citep{van2014accelerating}. Recently, however, \citet{linderman2019fast} developed a novel approximation -- FIt-SNE -- which reduces the asymptotic time complexity to $\mathcal {O}(N)$, enabling scaling to large data sets.

% We benchmark Python
Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.6.2). Benchmarks were performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original t-SNE parameters, as some implementations do not allow for their modification. To ensure a realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse brain data set~\citep{xxx}, containing 1.3 million single-cell gene-expression profiles. We preprocess the data using the standard single-cell analysis pipeline~\citep{xxx} and extract the top 50 principal components. To generate benchmark data sets of different sample sizes, we subsample the data set five times at each sample size. In total, we test each implementation on 30 different data sets.
